The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.tex
Prep for resubmission

  \usepackage{unicode-math}

@@

@@

\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}

@@

\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}

@@

\urlstyle{same} % disable monospaced font for URLs

@@

  pdftitle={A comparison of R tools for nonlinear least squares modeling},

@@

\title{A comparison of R tools for nonlinear least squares modeling}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2023-3-2}

@@

Our Google Summer of Code project ``Improvements to \texttt{nls()}''
considered the features and limitations of the R function \texttt{nls()}
with the aim of improving and rationalizing R tools for nonlinear
regression. The rich features of \texttt{nls()} are weakened by several
deficiencies and inconsistencies such as a lack of stabilization of the
Gauss-Newton solver. Further considerations are the usability and
maintainability of the code base that provides the functionality
\texttt{nls()} claims to offer. Various packages, including our
\texttt{nlsr}, provide alternative capabilities. We consider the
differences in goals, approaches, and features of different tools for
nonlinear least squares modeling in R. Discussion of these matters is
relevant to improving R generally as well as its nonlinear estimation
tools.

@@

\texttt{nls()} is the tool for estimating nonlinear statistical models
in base R, the primary software distribution from the Comprehensive R
Archive Network (\emph{\url{https://cran.r-project.org}}). The function
dates to the 1980s and the work related to Bates and Watts
(\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones that call
  calculations in other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound-constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to many pre-coded (\texttt{selfStart}) models that do not
  require initial parameter values.
\end{itemize}

@@

The tools we will consider are the base-R \texttt{nls()} function and
some packages that are available on the CRAN repository. Of these, we
will pay particular attention to \textbf{nlsr} (John C Nash and Duncan
Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
\textbf{minpack.lm} (Elzhov et al.
(\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau
(\protect\hyperlink{ref-gslnls23}{2023})) which are general nonlinear
least-squares solvers.

While we will provide capsule comments for some other CRAN packages, we
will not pursue those in the Bioconductor (Gentleman et al.
(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
repositories such as GitHub (\url{https://github.com}) and Gitlab
(\url{https://about.gitlab.com}).

>>>>

  \usepackage{unicode-math} % this also loads fontspec

@@

\ifPDFTeX\else
  % xetex/luatex font selection

@@

\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}

@@

\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}

@@

\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}

@@

\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}

@@

\urlstyle{same}

@@

  pdftitle={The Current State of R Tools for Nonlinear Least Squares Modeling},

@@

\title{The Current State of R Tools for Nonlinear Least Squares
Modeling}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2023-7-16}

@@

Our Google Summer of Code project ``Improvements to \texttt{nls()}''
considered the features and limitations of the R function \texttt{nls()}
with the aim of improving and rationalizing R tools for nonlinear
regression. The rich features of \texttt{nls()} are weakened by several
deficiencies and inconsistencies such as a lack of stabilization of the
Gauss-Newton solver. Further considerations are the usability and
maintainability of the code base that provides the functionality
\texttt{nls()} offers. Various packages, in particular \texttt{nlsr}
(John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative
capabilities. We consider the differences in goals, approaches, and
features of different tools for nonlinear least squares modeling in R.
Discussion of these matters is relevant to improving R generally as well
as its nonlinear estimation tools.

@@

\texttt{nls()} is the tool for estimating nonlinear statistical models
in base R, the primary software distribution from the Comprehensive R
Archive Network (CRAN). Users do not have to request that \texttt{nls()}
be loaded. The function dates to the 1980s and the work related to D. M.
Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) and others in
the S programming language.

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones that call
  calculations in other programming languages;
\item
  allows data to be weighted or subset;
\item
  can estimate bound-constrained parameters;
\item
  provides a mechanism for handling partially linear models;
\item
  permits parameters to be indexed over a set of related data;
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters;
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed;
\item
  links to many pre-coded (\texttt{selfStart}) models that do not
  require initial parameter values.
\end{itemize}

@@

Our Google Summer of Code project for 2021, ``Improvements to
\texttt{nls()}'' had the goal of providing a version of the code that
would address some of the deficiencies discovered over the long period
that \texttt{nls()} has been part of R. However, it was also critical to
preserve legacy functionality, and we did not achieve a result that
provided desired improvements and preserved legacy results
simultaneously. On the other hand, working with \texttt{nls()} suggested
ways package \texttt{nlsr} could be improved in features, while also
revealing how it has a very different internal structure and design.

The primary goal of this paper is to present some of these issues, with
a limited consideration of some other R tools that exist for nonlinear
least squares calculations. We hope to encourage others to engage,
possibly with us, to find ways to bring improvements and to consolidate
tools where reasonable. We want to ease the task of users in choosing
which tools to use. Unfortunately, it is simpler to create new packages
than to bring together and simplify existing ones.

\hypertarget{tools-considered}{%
\subsubsection{Tools considered}\label{tools-considered}}

Besides the base-R \texttt{nls()} function, we consider some packages
that are available in the CRAN repository. Of these, we will pay
particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr2023manual}{2023})), \textbf{minpack.lm}
(Elzhov et al. (\protect\hyperlink{ref-minpacklm12}{2012})), and
\textbf{gslnls} (Chau (\protect\hyperlink{ref-gslnls23}{2023})) which
are general nonlinear least-squares solvers. We will not pursue tools in
the Bioconductor (Gentleman et al.
(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
repositories such as GitHub and Gitlab.

The Hobbs weed infestation problem (John C. Nash
(\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
task that seems straightforward, but is quite nasty. Its very succinct
statement provides the ``short reproducible example'' much requested on
R mailing lists. This is a real problem from a field researcher, though
the units of measurement and background information are not available.
The data and graph follow.

\includegraphics{StateOfRToolsForNLS_files/figure-latex/ex01-1.pdf}

Let us consider there are \(n\) parameters and \(m\) residuals. The loss
function is \[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\] The gradient of
\(S(p)\) is \[ g = 2 * J' r\] where the Jacobian \(J\) is given by
elements \[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is
defined by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \] If we
expand the Hessian for nonlinear least squares problems, we find
\[  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this
expression. What is generally called \textbf{Newton's method} for
function minimization tries to set the gradient to zero (to find a
stationary point of the function \(S(p)\)). This leads to
\textbf{Newton's equation} \[ H \delta  = -g \] Given a set of
parameters \(p\), which provide \(H\) and \(g\), we solve this equation
for \(\delta\), adjust \(p\) to \(p + \delta\) and iterate, hopefully to
converge on a solution. Applying this to a sum of squares problem gives
\[ H \delta / 2 = (J'J + D) \delta = - J' r\] In this expression, only
the elements of \(D\) have second partial derivatives. Gauss, attempting
to model planetary orbits, had small residuals, and noted that these
multiplied the second partial derivatives of \(r\), so he approximated
\[ H / 2 \approx J' J\] by assuming \(D \approx 0\). This results in the
Gauss-Newton method where we solve \[ J' J \delta = - J' r\] though we
can avoid some loss of accuracy by not forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \] by one of several matrix decomposition
methods. On the other hand, when \(m >> n\), the matrix \(J' J\) uses
much less storage than \(J\).

In reality, there are many problems where \(D\) should not be ignored,
but the work to compute it precisely is considerable. Many work-arounds
have been proposed, of which the Levenberg-Marquardt stabilization
(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt
(\protect\hyperlink{ref-Marquardt1963}{1963})) is the most commonly
used. For convenience, we will use ``Marquardt'', as we believe he first
incorporated the ideas into a practical computer program.

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are not equivalent and are
deliberately crude choices. Workers performing many calculations of a
similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# model formulas}
\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3 }\SpecialCharTok{*}\NormalTok{ tt))}
\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{ c1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ c2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{c3 }\SpecialCharTok{*}\NormalTok{ tt))}
\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
\CommentTok{\#}
\CommentTok{\# Starting parameter sets}
\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =}\DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{)}
\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{1}\NormalTok{)}
\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

It is also possible to provide R functions for the residual and
Jacobian. \texttt{nls()} is not well-suited to using such a formulation,
but \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} work
specifically with such a formulation. This is usually more work for the
user if the formula setup is possible. To illustrate, we show the
functions for the unscaled 3 parameter logistic.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Logistic3U}
\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
  \CommentTok{\# This variant uses looping}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
\NormalTok{\}}

\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)}
\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ yy)}
\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy}
\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy }\SpecialCharTok{*}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt}
  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
\NormalTok{  jj}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Using a formula specification was a principal advantage made with
\texttt{nls()} when it became available in S sometime in the 1980s. It
uses a Gauss-Newton (i.e., unstabilized) iteration with a step reduction
line search (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})). This
works very efficiently as long as \(J\) is not ill-conditioned. Below we
see the default \texttt{nls()} algorithm does poorly on the example
problem. To save page space, we use 1-line result display functions from
package \texttt{nlsr}, namely \texttt{pnls()} and \texttt{pshort()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\end{Highlighting}
\end{Shaded}

Here we see the infamous ``singular gradient'' termination message of
\texttt{nls()}. Technically, it is the Jacobian that is singular, but R
does not distinguish scalar and vector functions, which yield partial
derivatives that are a gradient or Jacobian respectively.

\texttt{nls()} also offers the option of using the ``port'' library
algorithm choice (Fox, Hall, and Schryer
(\protect\hyperlink{ref-FoxHallSchryer1978}{1978})). There is a warning
in the \texttt{nls()} documentation that this ``appears unfinished'',
but it is successful in solving all three cases.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(unls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(snls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(tnls1p)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{print}\NormalTok{(unlx1) }
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
\end{Highlighting}
\end{Shaded}

Though we have found solutions, the Jacobian is essentially singular as
shown by its singular values. Note that these are \textbf{displayed} by
package \texttt{nlsr} in a single column in the output to provide a
compact layout, but the values do \textbf{not} correspond to the
individual parameters in whose row they appear; they are a property of
the whole problem.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(snlm1)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
\FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(gslnls)}
\NormalTok{ugslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
\FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
\FunctionTok{pnls}\NormalTok{(sgslnls1) }\CommentTok{\# Use summary() to get display}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
\FunctionTok{pnls}\NormalTok{(tgslnls1) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# Other algorithms}
\DocumentationTok{\#\# tgslnls1a\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))}
\DocumentationTok{\#\# pnls(tgslnls1a) }
\DocumentationTok{\#\# tgslnls1d\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))}
\DocumentationTok{\#\# pnls(tgslnls1d) }
\DocumentationTok{\#\# tgslnls1dd\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))}
\DocumentationTok{\#\# pnls(tgslnls1dd) }
\DocumentationTok{\#\# tgslnls1s2\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))}
\DocumentationTok{\#\# pnls(tgslnls1s2) }
\end{Highlighting}
\end{Shaded}

While default \texttt{gsl\_nls()} gives much of the same results as
\texttt{nlsLM()}, uncommenting the calls to the alternative algorithms
shows it can get the global minimum with any of these choices.

\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors
and singular values of the Jacobian (for diagnostic purposes). By
contrast, \texttt{minpack.lm::nlsLM()} and \texttt{nls()} use
\texttt{summary()}, which does not display the sum of squares, while
\texttt{print()} gives the sum of squares, but not the standard error of
the residuals. Such behaviour is consistent with other modeling
functions such as \texttt{lm()}.

The importance of the singular values is to allow us to gauge how
``nearly singular'' the Jacobian is at the solution, and the ratio of
the smallest to largest of the singular values is a simple but effective
measure. The ratios are 4.6641e-05 for Logistic3U, 0.021022 for
Logistic3S, and 0.001055 for Logistic3T; so Logistic3S is the ``least
singular''.

The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the
transformed model Logistic3T have a very large sum of squares, which may
suggest that these programs have failed. Since \texttt{nls()},
\texttt{nlsLM()}, and \texttt{gsl\_nls()} do not offer singular values,
we need to extract the Jacobian and compute its singular values. The
following script shows how to do this, using as Jacobian the
\texttt{gradient} element in the returned solution for these solvers.

\begin{verbatim}
## [1] 3.4641e+00 9.4495e-09 4.1324e-11
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{stspecial }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
\NormalTok{badstart }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stspecial, }\AttributeTok{data =}\NormalTok{ weeddf)}
\FunctionTok{print}\NormalTok{(badstart)}
\end{Highlighting}
\end{Shaded}

We also note that changing the start to

\begin{verbatim}
stt2 <- c(Asym=1, xmid=1, scal=2)
\end{verbatim}

allows both \texttt{nlsLM} and \texttt{gsl\_nls} to find the minimum.
\texttt{gsl\_nls} also succeeds when the particular form of the
Marquardt stabilization is changed from the default by the control
setting \texttt{scale\ =\ "levenberg"}.

We consider the code inside package \texttt{nlsr} to be competently
coded, with the advantage over \texttt{nls()} of the Levenberg-Marquardt
stabilization in common with \texttt{minpack.lm} and \texttt{glsnls}.
The success rate is high, but \texttt{nlsr} has its premature
terminations too. However, it provides diagnostic information in the
form of the singular values of the Jacobian that allows us to try
alternative starts, algorithms or programs in cases where there are
indications of early termination.

Let us write a slightly modified form of the nonlinear least squares
problem:

\[ S(p) = \sum_{i=1}^m { r_i(p)^2 }\]

R users are likely to think almost always that the functions \(r_i\) are
all of the same form, differing in the details of the data they use.
However, the famous Rosenbrock banana-shaped valley (Rosenbrock
(\protect\hyperlink{ref-Rosenbrock1960}{1960})) has
\[ r_1(p) = (1 - p_1)\] \[ r_2(p) = 10 (p_2 - {p_1}^2)\]

Thus, there can be problems where the idea of a ``model'' or a
``prediction'' makes no sense, though the nonlinear least squares
problem is still well-posed. In our code, we still refer to the
functions as ``residuals'', though this is purely a convenience.

We illustrate how to solve nonlinear least squares problems using a
function to define the residual. Note that \texttt{gsl\_nls()} requires
a vector \texttt{y} that is the expected value of what we have called
the residual, but in this case is actually the model. \texttt{gsl\_nls}
uses a numerical approximation for the Jacobian if the argument
\texttt{jac} is missing. Note functions \texttt{nlsr::pnls()} and
\texttt{nslr::pnlslm()} for a 1-line display of the results.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobnlfb }\OtherTok{\textless{}{-}} \FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{resfn =}\NormalTok{ hobbs.res, }\AttributeTok{jacfn =}\NormalTok{ hobbs.jac)}
\FunctionTok{pshort}\NormalTok{(hobnlfb) }\CommentTok{\# use print(hobnlfb) for more detail}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobnlm }\OtherTok{\textless{}{-}} \FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{jac =}\NormalTok{ hobbs.jac)}
\FunctionTok{pnlslm}\NormalTok{(hobnlm)  }
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobgsln }\OtherTok{\textless{}{-}} \FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{y =} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(hobgsln)}
\end{Highlighting}
\end{Shaded}
